Syntax and Semantics of Cedille

### 2020-12-23 docs/semantics/paper.tex
docs: semantics: rules and soundness proof
- update beta typing rule
- in soundness, fix cases for beta, rho, chi

    \infer{\Gamma\vdash [ t , t' ] \tpcheck \abs{\iota}{x}{T}{T'}}
          {\Gamma\vdash t \tpcheck T & \Gamma\vdash t' \tpcheck [t/x]T' & |t| =_{\beta\eta} |t'|} &
    \infer{\Gamma\vdash t.1 \tpsynth T}{\Gamma\vdash t \tpsynth \abs{\iota}{x}{T}{T'}} \\ \\
    \infer{\Gamma\vdash t.2 \tpsynth [t.1/x]T'}{\Gamma\vdash t \tpsynth \abs{\iota}{x}{T}{T'}} &

    \infer{\Gamma\vdash \beta\{t'\} \tpcheck \{ t \simeq t \}}{\Gamma\vdash \textit{FV}(t')\subseteq \textit{dom}(\Gamma)}  \\ \\    
    \infer{\Gamma\vdash \delta\ \mhyph\ t \tpcheck T}{\Gamma\vdash t\tpcheck \{ \absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}}  &
    \infer{\Gamma \vdash \rho\ t'\ @x.T\ \mhyph\ t \tpsynth [t_2/x]T}
          {
          \begin{array}{cc}
            \Gamma \vdash t' \tpsynth \{t_1 \simeq t_2\}
            & \quad \Gamma \vdash [t_2/x]T \tpsynth \star
            \\ \Gamma \vdash t \tpsynth T'
            & T' \cong [t_1/x]T
          \end{array}
          } \\ \\
    \infer{\Gamma \vdash \chi\ T\ \mhyph\ t\ \tpsynth T}
          {\Gamma \vdash T \tpsynth \star & \Gamma \vdash t \tpcheck T} &
    \infer{\Gamma\vdash \varphi\ t\ \mhyph\ t'\ \{t''\} \Leftrightarrow T}
          {\Gamma\vdash t\tpsynth \{t'\simeq t''\} & \Gamma\vdash t' \Leftrightarrow T}
  \end{array}
  \]
\caption{Rules for checking a term against a well-kinded type ($\Gamma \vdash t \tpcheck T$)
           and synthesizing a type for a term ($\Gamma \vdash t \tpsynth T$)}
\label{fig:tp}
\end{figure}

@@

We have modified the rules for equality types $\{ t \simeq t' \}$ so that we require
nothing of $t$ and $t'$ except that the set $\textit{dom}(\Gamma)$ of variables
declared by $\Gamma$ includes their free variables $\textit{FV}(t\ t')$.  Further modifications
over the version of CDLE in~\cite{stump18} are:
\begin{itemize}
\item To prove $\{ t \simeq t\}$, one now writes $\beta\{t'\}$,
  with the critical idea that $|\beta\{t'\}|$ erases to $|t'|$.  We call this the \textbf{Kleene trick},
  because it goes back to Kleene's numeric
  realizability~\cite{Kle65_Classical-Extensions-of-Intuitionistic-Mathematics},
  which accepts any number $n$ as a realizer 
  of a true equation.  Here, we accept any closed term $t$ as a realizer
  of $\{ t \simeq t\}$.  This means that in Cedille, any such term -- even otherwise untypable
  terms, non-normalizing terms, etc. -- have type $\{ t \simeq t\}$ for any term
  $t$.
\item The \(\rho\) construct allows one to rewrite occurrences of \(t_1\) to
  \(t_2\) in the synthesized type of the second subexpression.
  The version presented here requires a ``guide'' \(@x.T\), where \(x\)
  indicates the occurrences of \(t_1\) to substitute.
  \([t_2/x]T\) must be a well-kinded type, but it is not required that
  \([t_1/x]T\) also be, only that \(T'\) and \([t_1/x]T\) are convertible modulo
  erasure --- it is not guaranteed that \(t_1\) and \(t_2\) will have the same
  type, or will even be typeable.
  In Cedille, the guide is optional and the construct may be used to rewrite a
  contextually given type; a heuristic is used to produce a resulting type that
  is well-kinded whose details are beyond the scope of this document.
\item We adopt a strong form of Nuprl's \textbf{direct computation rules}~\cite{constable+86}:
  If we have a term $t'$ of type $T$ and a proof $t$ that $\{ t' \simeq t''\}$, then we may conclude that
  $t''$ has type $T$ by writing the annotated term $\varphi\ t\ \mhyph\ t'\ \{t''\}$, which
  erases to $t''$.
\item Where the previous version of CDLE uses $\beta$-equivalence for (erased) terms, we here adopt $\beta\eta$-equivalence.  This
  allows us to observe in many cases that retyping functions are actually $\beta\eta$-equivalent to $\absu{\lambda}{x}{x}$.
  While $\beta\eta$-equivalence takes more work to incorporate into intrinsic 
  type theory~\cite{geuvers92},
  it raises no difficulties for our extrinsic one.
\item In this version, we add an explicit axiom $\delta$ saying that Church-encoded boolean \emph{true} is different from
  \emph{false}.
  In the implementation of Cedille, this is generalized to the rule where the
  proof \(t\) synthesizes an equation \(\{t_1 \simeq t_2\}\) in which \(|t_1|\)
  and \(|t_2|\) are separable using the \emph{B\"ohm-out algorithm}~\cite{BDPR79_Bohm-Algorithm}.

@@

\startcase{.2cm}
\[
    \infer{\Gamma\vdash \beta\{t'\} \tpcheck \{ t \simeq t \}}{\Gamma\vdash \textit{FV}(t)\subseteq \textit{dom}(\Gamma)}  
\]
$[\sigma|t'|]_{\cbe}\in\interp{\{ t \simeq t \}}_{\sigma,\rho}$ follows directly from the semantics of equality types.

@@

\startcase{.2cm}
\[
    \infer{\Gamma\vdash \rho\ t'\ \mhyph\ t \Leftrightarrow [t_2/x]T}
          {\Gamma\vdash t' \tpsynth t_1 \simeq t_2 & \Gamma \vdash t \Leftrightarrow [t_1/x]T} 
\]
By the IH applied to the first premise, $\sigma|t_1| =_{\beta\eta} \sigma|t_2|$.  The result then follows by the IH applied to the second premise,
and Lemma~\ref{lem:termsubstinterp}.

\startcase{.2cm}
\[
    \infer{\Gamma\vdash \chi\ T\ \mhyph\ t \tpcheck T'}
          {\Gamma\vdash T\tpsynth \star & \Gamma\vdash t \tpcheck T & T \cong T'} 
\]
Using the IH for the first premise and the assumption that $\interp{T'}_{\sigma,\rho}$ is in $\mathcal{R}$ and hence defined,
we can apply the IH to the third premise to get $\interp{T}_{\sigma,\rho} = \interp{T'}_{\sigma,\rho}$.  Using this and the IH for second premise,
we get the desired conclusion, using also the definition of erasure.

\startcase{.2cm}
\[
    \infer{\Gamma\vdash \chi\ T\ \mhyph\ t \tpsynth T}
          {\Gamma\vdash T\tpsynth \star & \Gamma\vdash t \tpsynth T' & T \cong T'} 
\]
By the IH applied to the second premise, we have $[\sigma|t|]_{\cbe}\in\interp{T'}_{\sigma,\rho}\in\mathcal{R}$.
Using definedness of $\interp{T'}_{\sigma,\rho}$ and the IH applied to the first premise, we can apply the
IH to the third premise to get $\interp{T}_{\sigma,\rho} = \interp{T'}_{\sigma,\rho}$, from which the desired
conclusion follows by definition of erasure.

>>>>

    \infer{\Gamma\vdash [ t , t' ] \tpcheck \abs{\iota}{x}{T}{T'}}
          {\Gamma\vdash t \tpcheck T & \Gamma\vdash t' \tpcheck [t/x]T' & |t| =_{\beta\eta} |t'|} &
    \infer{\Gamma\vdash t.1 \tpsynth T}{\Gamma\vdash t \tpsynth \abs{\iota}{x}{T}{T'}} \\ \\
    \infer{\Gamma\vdash t.2 \tpsynth [t.1/x]T'}{\Gamma\vdash t \tpsynth \abs{\iota}{x}{T}{T'}} &
    \infer{\Gamma\vdash \beta\{t'\} \tpcheck \{ t_1 \simeq t_2 \}}
          {\Gamma\vdash \textit{FV}(t'\ t_1\ t_2)\subseteq \textit{dom}(\Gamma) \quad |t_1| =_{\beta\eta} |t_2|}  \\ \\    
    \infer{\Gamma\vdash \delta\ \mhyph\ t \tpcheck T}{\Gamma\vdash t\tpcheck \{ \absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}}  &
    \infer{\Gamma \vdash \rho\ t'\ @x.T\ \mhyph\ t \tpsynth [t_2/x]T}
          {
          \begin{array}{cc}
            \Gamma \vdash t' \tpsynth \{t_1 \simeq t_2\}
            & \quad \Gamma \vdash [t_2/x]T \tpsynth \star
            \\ \Gamma \vdash t \tpsynth T'
            & T' \cong [t_1/x]T
          \end{array}
          } \\ \\
    \infer{\Gamma \vdash \chi\ T\ \mhyph\ t\ \tpsynth T}
          {\Gamma \vdash T \tpsynth \star & \Gamma \vdash t \tpcheck T} &
    \infer{\Gamma\vdash \varphi\ t\ \mhyph\ t'\ \{t''\} \Leftrightarrow T}
          {\Gamma\vdash t\tpsynth \{t'\simeq t''\} & \Gamma\vdash t' \Leftrightarrow T}
  \end{array}
  \]
\caption{Rules for checking a term against a well-kinded type ($\Gamma \vdash t \tpcheck T$)
           and synthesizing a type for a term ($\Gamma \vdash t \tpsynth T$)}
\label{fig:tp}
\end{figure}

@@

We have modified the rules for equality types $\{ t \simeq t' \}$ so that we require
nothing of $t$ and $t'$ except that the set $\textit{dom}(\Gamma)$ of variables
declared by $\Gamma$ includes their free variables $\textit{FV}(t\ t')$.  Further modifications
over the version of CDLE in~\cite{stump18} are:
\begin{itemize}
\item To prove $\{ t_1 \simeq t_2\}$ for definitionally equal terms (that is,
  terms that are \(\beta\eta\)-equivalent modulo erasure), one now writes $\beta\{t'\}$,
  with the critical idea that $|\beta\{t'\}|$ erases to (a possibly unrelated) $|t'|$.
  We call this the \textbf{Kleene trick}, because it goes back to Kleene's numeric
  realizability~\cite{Kle65_Classical-Extensions-of-Intuitionistic-Mathematics},
  which accepts any number $n$ as a realizer 
  of a true equation.  Here, we accept any term $t'$ as a realizer of $\{ t_1
  \simeq t_2\}$, provided the free variables of \(t'\) are declared in the
  context.
  This means that in Cedille, any such term --- even otherwise untypable
  terms, non-normalizing terms, etc. --- prove trivially true equations.
\item The \(\rho\) construct allows one to rewrite occurrences of \(t_1\) to
  \(t_2\) in the synthesized type of the second subexpression.
  The version presented here requires a ``guide'' \(@x.T\), where \(x\)
  indicates the occurrences of \(t_1\) to substitute.
  \([t_2/x]T\) must be a well-kinded type, but it is not required that
  \([t_1/x]T\) also be, only that \(T'\) and \([t_1/x]T\) are convertible modulo
  erasure --- it is not guaranteed that \(t_1\) and \(t_2\) will have the same
  type, or will even be typeable.
  In Cedille, the guide is optional and the construct may be used to rewrite a
  contextually given type; a heuristic is used to produce a resulting type that
  is well-kinded whose details are beyond the scope of this document.
\item We adopt a strong form of Nuprl's \textbf{direct computation rules}~\cite{constable+86}:
  If we have a term $t'$ of type $T$ and a proof $t$ that $\{ t' \simeq t''\}$, then we may conclude that
  $t''$ has type $T$ by writing the annotated term $\varphi\ t\ \mhyph\ t'\ \{t''\}$, which
  erases to $t''$.
\item Where the previous version of CDLE uses $\beta$-equivalence for (erased) terms, we here adopt $\beta\eta$-equivalence.  This
  allows us to observe in many cases that retyping functions are actually $\beta\eta$-equivalent to $\absu{\lambda}{x}{x}$.
  While $\beta\eta$-equivalence takes more work to incorporate into intrinsic 
  type theory~\cite{geuvers92},
  it raises no difficulties for our extrinsic one.
\item In this version, we add an explicit axiom $\delta$ saying that Church-encoded boolean \emph{true} is different from
  \emph{false}.
  In the implementation of Cedille, this is generalized to the rule where the
  proof \(t\) synthesizes an equation \(\{t_1 \simeq t_2\}\) in which \(|t_1|\)
  and \(|t_2|\) are separable using the \emph{B\"ohm-out algorithm}~\cite{BDPR79_Bohm-Algorithm}.

@@

\startcase{.2cm}
\[
  \infer{\Gamma\vdash \beta\{t'\} \tpcheck \{ t_1 \simeq t_2 \}}
  {\Gamma\vdash \textit{FV}(t'\ t_1\ t_2)\subseteq \textit{dom}(\Gamma) \quad |t_1| =_{\beta\eta} |t_2|}
\]
Since by assumption \(|t_1| =_{\beta\eta} |t_2|\) it follows that \(\sigma|t_1|
=_{\beta\eta} \sigma|t_2|\), and also by assumption it follows that
\(\textit{FV}(t_1\ t_2) \subseteq \textit{dom}(\sigma)\).
Also, we see \(\sigma|t'|\) is a closed term.
So, $[\sigma|t'|]_{\cbe}\in\interp{\{ t_1 \simeq t_2 \}}_{\sigma,\rho}$ follows
directly from the semantics of equality types.

@@

\startcase{.2cm}
\[
  \infer{\Gamma \vdash \rho\ t'\ @x.T\ \mhyph\ t \tpsynth [t_2/x]T}
  {
    \begin{array}{cc}
      \Gamma \vdash t' \tpsynth \{t_1 \simeq t_2\}
      & \quad \Gamma \vdash [t_2/x]T \tpsynth \star
      \\ \Gamma \vdash t \tpsynth T'
      & T' \cong [t_1/x]T
    \end{array}
  }
\]
By the IH applied to the first premise in the first row, $\sigma|t_1| =_{\beta\eta}
\sigma|t_2|$.
By the IH applied to the first premise in the second row,
\([\sigma|t|]_{c\beta\eta} \in \interp{T'}_{\sigma.\rho}\).
By the IH applied to the second premise in the second row,
\(\interp{T'}_{\sigma,\rho} = \interp{[t_1/x]T}_{\sigma,\rho}\).
The result then follows by applying Lemma~\ref{lem:termsubstinterp}.

\startcase{.2cm}
\[
  \infer{\Gamma \vdash \chi\ T\ \mhyph\ t\ \tpsynth T}
  {\Gamma \vdash T \tpsynth \star & \Gamma \vdash t \tpcheck T}
\]
We apply the IH for the first premise to get that
$\interp{T}_{\sigma,\rho}$ is in $\mathcal{R}$ and hence defined.
Using this, we apply the IH on the second premise to get \([\sigma|t|]_{\cbe}
\in \interp{T}_{\sigma,\rho}\), which by the definition of erasure is what we
must show.
